---
title: "Practice Assignment"
author: "Jonathan Dale"
date: "2026-02-09"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Instructions:** Create a ***pdf*** file with all relevant code and outputs using **Rmarkdown**. Do not submit any other file formats and do not print the entire dataset with the answers. *ONLY include relevant outputs.*

### Q1) Use the 'carseats.csv' for the data set. Read the description of the dataset and answer the following questions.

a)  Import the "carseats" dataset, look at the first few rows and inspect the data types of the variables in the dataframe.
b)  Change the variables "ShelveLoc, urban, US" into a factor variables.
c)  create a new variable called "profit" which stands for "Income - Advertising"
d)  Check for missing data. If you have missing data remove the corresponding rows from the dataset.
e)  How many "Good" shelving locations are there in the dataset?
f)  How many stores are inside the USA? create a separate data frame containing all stores from USA. Name the data set as "stores_USA"
g)  create another data set called "HighUrban_USSales" using 'stores_USA' data set. Where, sales are greater than 7 thousand and stores are located in Urban areas.
h)  Remove "US" and "Urban" columns from the "HighUrban_USSales" dataset.
i)  For one the above subset, write to a new CSV file

---
```{r echo=TRUE}
library(readr)
## a) Import the "carseats" dataset, look at the first few rows and inspect the data types of the variables in the dataframe
carseats_1 <- read.csv("~/STATS/STAT-515-002/homework/hw-1/carseats-1.csv")
str(carseats_1)
head(carseats_1)

## b)  Change the variables "ShelveLoc, urban, US" into a factor variables.

    # View the structure first
  str(carseats_1)
    # Adjust the variables, convert into factor variable type
  carseats_1$ShelveLoc = factor(carseats_1$ShelveLoc)
  carseats_1$Urban = factor(carseats_1$Urban)
  carseats_1$US = factor(carseats_1$US)
    # Verify that the variables were converted into factor variables
  str(carseats_1)


# c)  create a new variable called "profit" which stands for "Income - Advertising"
  
    # Create the new variable
  profit = carseats_1$Income - carseats_1$Advertising
    # Then add the new variable to the data set
  carseats_1$profit = profit
    # verify
  str(carseats_1)

# d)  Check for missing data. If you have missing data remove the corresponding rows from the dataset.
  
    # Do we have missing data?
  table(is.na(carseats_1)) ## Yes we have missing data!
    # Create new dataset by removing the missing values from the original.
  carseats_clean = na.omit(carseats_1)


# e)  How many "Good" shelving locations are there in the dataset?
    
    # Take a subset with only the ShelveLoc = "Good"
  Good_location = subset(carseats_1, ShelveLoc == "Good")
    # Use nrow() to count the new subset
  nrow(Good_location)

    # Alternatively, just apply a filter and wrap with the nrow() function.
  nrow(subset(carseats_1, ShelveLoc == "Good"))


# f)  How many stores are inside the USA? create a separate data frame containing all stores from USA. Name the data set as "stores_USA"

    # Create new dataframe as a subset of original
  stores_USA = subset(carseats_1, US == "Yes")
    # use nrow() function to count rows of the subset where US == "Yes"
  nrow(stores_USA)


# g)  create another data set called "HighUrban_USSales" using 'stores_USA' data set. Where, sales are greater than 7 thousand and stores are located in Urban areas.
  
    # Create new data frame
  HighUrban_USSales = subset(stores_USA, Sales > 7 & Urban == "Yes")
  

# h)  Remove "US" and "Urban" columns from the "HighUrban_USSales" dataset.
  
    # Check the original structure, remove the columns, verify  
  str(HighUrban_USSales)
  HighUrban_USSales = subset(HighUrban_USSales, select = -c(US, Urban))
  str(HighUrban_USSales)



# i)  For one the above subset, write to a new CSV file
    ## Write to file removing the row column
  write.csv(HighUrban_USSales, "HighUrban_USSales.csv", row.names = F)


```




### Q2) See the following code of a function and explain what it does. Suggest a suitable name for the function and rename. Demonstrate how the function works when you have numerical data and character data.

The code below defines a function named 'function1' that takes a single argument x. If the vector, x, has length 1 or less, the function returns NULL. If length greater than 1, the function returns a new vector with all the elements except the last one. 
I would re-name this to remove_last_element() to better describe what it dose. 

```{r,eval=FALSE}
function1 <- function(x) {
  if (length(x) <= 1) return(NULL)
  x[-length(x)]
}
```

### Q3) Write a function to compute the sample variance of a numerical vector. Use the equation of the variance to write the function.

```{r}
sample_variance = function(x) {
  # Validate the input
  if (! is.numeric(x)) {
    stop("Must be numeric vector")
  }
  if (length(x) < 2) {
    stop("Input vector must have at least 2 elements to compute sample variance.")
  }
  # Input validation sources from Google AI overview.
  
  # Calculate number of observations, mean, sum of squared differences, and variance
  n <- length(x)
  xbar <- mean(x, na.rm = TRUE)
  sum_sq_diff = sum((x - xbar)^2)
  sample_variance = sum_sq_diff / (n -1)
  return(sample_variance)
}


my_sample_vector = c(10, 12, 15, 11, 14, 13) # Create a sample vector
my_sample_variance_function_call = sample_variance(my_sample_vector) # Call the function, store results in variable
cat("Manual variance function:", my_sample_variance_function_call, "\n" )

## Lets verify the results with the built in function...
built_in_sample_variance = var(my_sample_vector)
cat("Built in sample variance var():", built_in_sample_variance, "\n")


```



