---
title: "Homework 3"
author: "Jonathan Dale"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    code_folding: show   # all chunks shown by default, can change to hide
  pdf_document: default
---
```{css, echo=FALSE}

<!-- define CSS rules to limit the height of code blocks -->

pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 900px;
}

.scroll-100 {
  max-height: 300px;
  overflow-y: auto;
  background-color: inherit;
}
/*
  Source https://bookdown.org/yihui/rmarkdown-cookbook/html-scroll.html
*/
```

Instructions: Include all codes and R-outputs that are necessary for each question. Submission: You MUST submit a PDF file.

#### Question 1: Regression Models:

Is there a relationship between student math test scores and socioeconomic variables? The data set `CASchools.csv` contains data on test performance, school demographics, and student demographic background for school districts in California. Remove the variable `county` before the analysis.

Please use the Data set description document to learn more about the data set.


a). Fit a linear regression model to predict math test scores using the variables in the dataset. Discuss your results, making sure to cover the following points:

1.  What do you observe about the relationship between these predictors and math test scores?

2.  Are there any insignificant variables?

3.  Which predictor is explaining math scores the best?

4.  Explain the interpretation of the coefficient on lunch in terms of the response variable.

```{r question 1.a, message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)

data = read.csv("CASchools.csv")
data_1a = data
glimpse(data_1a)

# Data analysis and cleaning:
# Remove the "county" column
# The "district" and "school" are ID attributes which are not useful for regression analysis,
# they are removed from the data and will not be use in the model.
# Two columns are character type, "computer" & "income", convert to numeric.
# The "grades" column is also character type, convert to factor for analysis.
data_1a = read.csv("CASchools.csv") |>
  select(-county, -district, -school) |>
  mutate(
    across(
      c(students, computer, income),
      ~ suppressWarnings(as.numeric(.x))
    ),
    grades = factor(grades)
  )

glimpse(data_1a)



# Multiple linear regression (ISLR2 §3.6.3)
# Source: ISLRv2 https://hastie.su.domains/ISLR2/ISLRv2_corrected_June_2023.pdf.download.html
fit1 = lm(math ~ ., data = data_1a)


## To see a list of hidden values use 'str(summary(model))' 
## https://stackoverflow.com/a/5588638
# Extract just the p-values from the model (excluding the intercept)
# Filter for Insignificant variables where p > 0.05
p_values = summary(fit1)$coefficients[, "Pr(>|t|)"][-1]
insignificant_vars = p_values[p_values > 0.05]
print(insignificant_vars)



# Find the strongest predictor with scaling of all the variables
# Create a new model using scaled variables.
data_1a_scaled = data_1a |>
  mutate(across(where(is.numeric), scale))

# Fit a new model with scaled values
fit1_scaled = lm(math ~ ., data_1a_scaled)


# Extract coefficients (excluding the intercept) to find the strongest predictor
betas = coef(fit1_scaled)[-1]
strongest_beta_name = names(which.max(abs(betas)))
strongest_beta_value = betas[strongest_beta_name]

c("Strongest Predictor:", strongest_beta_name)


# Remove the insignificant variables to find largest |t|
data_1a_significant = data_1a_scaled |> select(-names(insignificant_vars))
names(data_1a_significant)
fit1_sig = lm(math ~ ., data_1a_significant)

t_values_std = summary(fit1_sig)$coefficients[, "t value"][-1]
max_t_name = names(which.max(abs(t_values_std)))
max_t_value = t_values_std[max_t_name]

c("Max |t| value:", round(max_t_value, 4))





# # Create a pairwise correlation plot with the pairs function using only numeric columns
# # scatter plot matrix to observe the behavior of each variable
library(psych)
pairs.panels(data_1a |> select(where(is.numeric)))


# Simple regression models  for comparison:
# math ~ lunch
fit_lunch = lm(math ~ lunch, data = data_1a)
math_vs_lunch = c(r_squared = summary(fit_lunch)$r.squared, residual_se = summary(fit_lunch)$sigma)
# cor.test(data_1a$math, data_1a$lunch, alternative = 'two.sided')

# math ~ income
fit_income = lm(math ~ income, data = data_1a)
math_vs_income = c(r_squared = summary(fit_income)$r.squared, residual_se = summary(fit_income)$sigma)
# cor.test(data_1a$math, data_1a$income, alternative = 'two.sided')


# Extract r-squared and residual SE (sigma) from the model
original_model = c(r_squared = summary(fit1)$r.squared, residual_se = summary(fit1)$sigma)

## Full regression vs simple regression comparison.
comparison_output_1a <- c(
  "========================================",
  "Original model (full regression)",
  capture.output(print(original_model)),
  "",
  "Math vs lunch alone",
  capture.output(print(math_vs_lunch)),
  "",
  "Math vs income alone",
  capture.output(print(math_vs_income)),
  "",
  "We can see from the results above, neither lunch or income create a better model than the original.",
  "========================================"
)
cat(paste(comparison_output_1a, collapse = "\n"), "\n")


```

**Discussion (Question 1a).**

1. **Predictors vs math.** The full regression model shows strong negative association with `lunch` (Percent qualifying for reduced-price lunch), and a smaller positive association with `income` and math both clearly displayed in the plot matrix. The `expenditure`, `students`, `teachers`, and `calworks` are weaker in this fit after adjusting for the others.

2. **Insignificant values (p > 0.05).** I used a filter to sort parameters where p > 0.05, this means the parameter is insignificant and will not contribute anything meaningful to the model. Store these values inside a variable named `insignificant_vars`. The following variables have p-values above 0.05: `students`, `teachers`, `calworks`, `expenditure`; also `computer` is marginal though still above the threshold.

3. **Strongest predictor.** By evidence in the data, it is clear that `lunch` has the largest |t| and smallest p-value. among the predictors making it the strongest linear signal in this model. This is found by scaling all the parameters in the data set.

4. **`lunch` coefficient.** With other predictors held fixed, each **one percentage-point** increase in `lunch` is associated with an estimated **decrease** of about **0.37** points in average `math` test score. In other words, areas with higher percent of students qualifying for reduced-price lunch, our model predicts a significantly larger portion of students with lower average math scores than areas with lower percentage of students on reduced-prices lunch.


b). Do you think you can fit a better model than the model in part (a)? Fit a new model that you think is a better fit for the data. What do you observe about this new model? (Use an appropriate variable selection algorithm and use cross-validation to validate your answers).

```{r question 1.b, message=FALSE, warning=FALSE}

model_full = lm(math ~ ., data = data_1a)
model_step = step(model_full, direction = "both")
summary(model_step)

library(caret)
set.seed(335)
ctrl = trainControl(method = "cv", number = 10)
cv_model = train(math ~ .,
                 data = data_1a,
                 method = "lm",
                 trControl = ctrl,
                 na.action = na.omit)

# Find the cross-validated MSE
cv_model$results$RMSE^2

# Full anova analysis
anova(fit1, model_step)

# Extract r-squared and residual SE (sigma) from the new model
step_model = c(r_squared = summary(model_step)$r.squared, residual_se = summary(model_step)$sigma)


## Full regression (part 1a) vs VARIABLE_SELECTION comparison.
comparison_output_1b <- c(
  "========================================",
  "Original model (full regression)",
  capture.output(print(original_model)),
  "",
  "New model, stepwise selection",
  capture.output(print(step_model)),
  "",
  "New stepwise selection model part anov",
  capture.output(print(anova(fit1, model_step))),
  "",
  "We can see from the results above that the new model preforms slightly better than the original model",
  "========================================"
)
cat(paste(comparison_output_1b, collapse = "\n"), "\n")


```

**1.b** By using Variable Selections methods we can choose a better model. We use best subset selection, 



c). Write the equation for the least square fit from part b).

```{r}

```

d). Compare the R2 and Testing MSE for models from part (a) and (b).

```{r}

```

e). Using a residuals plot for part a and part b models, state the appropriateness of using a linear model for this problem.

f). Incorporate an interaction term into your multiple linear regression and respond to the following:
<!-- ISLR book, page 127, 3.6.4 Interaction Terms -->

1.  Provide analytical justification for your interaction term.

2.  Compare your results with the models from parts a) and b).

g). Fit a 'Ridge Regression' Model for the dataset and interpret the results.

h). Fit a 'LASSO Regression' model for the dataset and interpret the results.

i). Compare all Testing MSE values that you got from parts a,b,f,g and h. Which model do you recommend?




#### Problem 2: Tree based models:

This question should be answered using the ***"Wine Quality"*** data set. Description about the data set can be found on the link provided. Objective of this question is to fit an regression tree model to predict quality of wine.

a).  Produce some numerical and graphical summaries of the data set. Explain the relationships.

b).  Create a training set with 80% of the observations, and a testing set containing the remaining 20%.

c).  Fit a regression tree with *quality* as the response variable using the training set. Plot the tree and interpret the results. What test MSE do you obtain?

d).  Use cross-validation in order to determine the optimal level of tree complexity. Does pruning the tree improve the test MSE?

e).  Use random forests to analyze this data. What test MSE do you obtain?

f).  Use the *importance()* function to determine which variables are most important.












